---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "January 25, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

https://github.com/crj688/LastTime

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data. **You do NOT need to rerun the census_data_prep.R file.** I'm
including it and encouraging you to take a look through it to get a sense of what a
typical data prep script might look like. We'll be learning to do all of the things
shown in that file.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
 county_df <- read.table("data/county.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
 county_MI <- read.table("data/county_MI.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
```

*What does each row represent?*
Each row represents a county.

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. 

*Are there any NA values?* 
Yes there are NA Values. 1 for Income, IncomeErr, and Child Poverty in County_DF.
None in County_MI

*What differences do you notice between the structure of the two data frames?*
The first one is much larger. Also second one just contains one state, MI. vs all states and it has a peninsula field.

```{r}
summary(county_df)
summary(county_MI)
```

The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

**Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.**

### Problem 2 - Basic plots for median Income

Note that the field called `Income` is the median income
for that county. You do **NOT** need to compute a median. The values are the medians.

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 

*Write a short paragraph describing the distribution of median incomes.*
The median income is around 45,000 for the entire country, USA. This is stated in the summary and it shows in the distribution amongst all of the plots below. The outlier incomes are 4th quartile. Over 500 counties have an average income of 45,000.

```{r}
ggplot(county_df) + geom_histogram(aes(Income), fill="#FF9999", colour="black")
ggplot(county_df) + geom_boxplot(aes(x= "USA", y=Income, fill="#FF9999"))
ggplot(county_df) + geom_density(aes(Income), fill="#FF9999", colour="black")
ggplot(county_df) + geom_violin(aes(x= "USA", y=Income, fill="#FF9999"))
```

### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US. 

```{r histo_totpop}
ggplot(county_df) + geom_histogram(aes(TotalPop), fill="#FF9999", colour="black")
```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
ggplot(county_df) + geom_histogram(aes(TotalPop), fill="#FF9999", colour="black") + 
                    ggtitle("Distribution of population for US counties") +
                    xlab("Total Population") +
                    ylab("Number of counties")
```

Repeat the histogram but only for counties with less than 1.5 million people. Do NOT use
dplyr to do the filtering - instead, use boolean indexing (see intro to dataframes Rmd file).

```{r histo_lt1.5m}

ggplot(county_df[county_df$TotalPop < 1500000,], aes(x=TotalPop, fill="#FF9999")) +                           geom_histogram() + ggtitle("Distribution of population for US counties") +
                                      xlab("Total Population") +
                                      ylab("Number of counties") 

```


**Hacker Extra** - Create a function that uses the
[Freedman-Diaconis](https://en.wikipedia.org/wiki/Freedman%E2%80%93Diaconis_rule) rule for calculating bin width and number of bins. The function should take a vector (e.g. a dataframe column) as its
input and return the integer number of bins recommended for the histogram.

```{r}
numbins_FD <- function(data_col){
  
}
```



### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped violin plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

*Do the peninsulas differ with respect to distribution of median income over the counties?*
Both Peninsulas main distribution of county median income is around 40,000. The Lower Peninsula has a broader distribution than the Upper Peninsula but the bulk of counties have an income slightly higher than 40,000 while the with the distribution for the Upper Peninsula, the bulk of counties income is slightly lower than 40,000.
Both Pennisulas, Michigan all together have a slightly lower median income than the country as a whole.

```{r vp_inc_by_peninsula}
ggplot(county_MI) + geom_violin(aes(x = peninsula, y = Income, fill="#FF9999")) +
                 theme(axis.title.x = element_text(face="bold", colour="steelblue", size=20), axis.text.x  = element_text(angle=0, hjust =.50, size=10)) +
                theme(axis.title.y = element_text(face="bold", colour="steelblue", size=20), axis.text.y  = element_text(angle=0, hjust=.50, size=10)) +
ggtitle("Distribution of Income by Peninsula")
           

```

Create several boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r boxplots}
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Men, fill="#FF9999")) +
                 theme(axis.title.x = element_text(face="bold", colour="steelblue", size=20), axis.text.x  = element_text(angle=0, hjust =.50, size=10)) +
                theme(axis.title.y = element_text(face="bold", colour="steelblue", size=20), axis.text.y  = element_text(angle=0, hjust=.50, size=10)) +
ggtitle("Distribution of Men by Peninsula")

ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Women, fill="#FF9999")) +
                 theme(axis.title.x = element_text(face="bold", colour="steelblue", size=20), axis.text.x  = element_text(angle=0, hjust =.50, size=10)) +
                theme(axis.title.y = element_text(face="bold", colour="steelblue", size=20), axis.text.y  = element_text(angle=0, hjust=.50, size=10)) +
ggtitle("Distribution of Women by Peninsula")

ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Black, fill="#FF9999")) +
                 theme(axis.title.x = element_text(face="bold", colour="steelblue", size=20), axis.text.x  = element_text(angle=0, hjust =.50, size=10)) +
                theme(axis.title.y = element_text(face="bold", colour="steelblue", size=20), axis.text.y  = element_text(angle=0, hjust=.50, size=10)) +
ggtitle("Distribution of African Americans by Peninsula")

ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Citizen, fill="#FF9999")) +
                 theme(axis.title.x = element_text(face="bold", colour="steelblue", size=20), axis.text.x  = element_text(angle=0, hjust =.50, size=10)) +
                theme(axis.title.y = element_text(face="bold", colour="steelblue", size=20), axis.text.y  = element_text(angle=0, hjust=.50, size=10)) +
ggtitle("Distribution of Citizens by Peninsula") 

```
*Distribution of Men by Peninsula*
In the plot Distribution of Men by Peninsula, you can see that there is obviously more men in the Lower Peninsula. The distribution is greater. There are even some counties with outlier numbers of men in them. A lot of these outliers occur in the 4th Quartile. This plot shows a county with the max of about 800,000 men. In the Upper Peninsula the median number of men per county is lower than the Lower Peninsula. The distribution there is a lot closer together. The county with the most men looks like it is a max number of about 20,000/30,000.

*Distribution of Women by Peninsula*
In the plot Distribution of Women by Peninsula, you can see that there is obviously more Women in the Lower Peninsula. The distribution is greater. There are even some counties with outlier numbers of women in them. A lot of these outliers occur in the 4th Quartile. This plot shows a county with the max of 900,000 women. In the Upper Peninsula the median number of Women per county is lower than the Lower Peninsula. The distribution there is a lot closer together. The county with the most women looks like it is a max number of about 30,000/40,000.

*Distribution of Blacks by Peninsula*
In the plot Distribution of African Americans by Peninsula, you can see that there is obviously more Blacks in the Lower Peninsula. The distribution is greater. There are even some counties with outlier percentages of Blacks in them. A lot of these outliers occur in the 4th Quartile. This plot shows a county with a max percent of about 40% Blacks. In the Upper Peninsula the median percentage of Blacks per county is lower than the Lower Peninsula. The distribution there is a lot closer together. The county with the most women looks like it is a max percentage of about 18%.

*Distribution of Citizens by Peninsula*
In the plot Distribution of Citizens by Peninsula, you can see that there are more Citizens in the Lower Peninsula. The distribution is greater. There are even some counties with outlier number of Citizens in them. A lot of these outliers occur in the 4th Quartile. This plot shows a counties with a max percent of about 1 million Citizens in them. In the Upper Peninsula the median number of Citizens per county is lower than the Lower Peninsula. The distribution there is a lot closer together. The county with the most Citizens in it looks like it is a max of about 30,000.

### Problem 5

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how unemployment rate compares by region.  Make sure your graph has a title and good axis
labels. 

```{r faceted_histo}
ggplot(data = county_df) + geom_histogram(aes(Unemployment), fill="#FF9999", colour="black")                  +  ggtitle("Distribution of Unemployment for US Counties by Region") +
                    xlab("Unemployment Percentage") +
                    ylab("Number of counties") +
      facet_wrap(~Region)
```

Now do a group of boxplots of mean commute time by region.

```{r bp_commute_region}
ggplot(data = county_df) +  geom_boxplot(aes(x= "Average Commute Time", y = MeanCommute,   
                                         fill="#FF9999")) +                 
                ggtitle("Distribution of Average Commute Time for Counties by Region") +
                xlab("Average Commute Time") +
                ylab("Number of counties") +
      facet_wrap(~Region)
  
  
  
```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Division in the US. Make sure
your graph has a title and good axis labels. Figure out how to format the
x-axis labels so that they don't obscure each other.

```{r bar_counties_by_division}
ggplot(county_df) + geom_bar(aes(x = Division, stat = "count", fill="#FF9999")) +
                 theme(axis.title.x = element_text(face="bold", colour="steelblue", size=20),                  axis.text.x  = element_text(angle=90, hjust =.50, size=10)) +
                 theme(axis.title.y = element_text(face="bold", colour="steelblue", size=20),                 axis.text.y  = element_text(angle=0, hjust=.50, size=10)) +
ggtitle("Counties by Division")


```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}
ggplot(county_MI) + 
  geom_point(aes(x=Income, y=MeanCommute))

```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula and the size is mapped to total population.

```{r xy_inc_commute_peninsula}
ggplot(county_MI) + 
  geom_point(aes(x=Income, y=MeanCommute, colour=peninsula, size=TotalPop))
```

*Your thoughts on the relationship between these variables?*
Well it seems as in counties with a higher average income people have a longer average commute time to work. It could be that counties with a lower average income people may make use of public transportation or have non relaible transportaion, therefore people tend to take jobs that are closer to home. One of the counties with the one of the largest poopulations of at least 1.5 million in Michigan has average income of about 43,000 and an average commute of 25 minutes to work.

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

```{r xy_faceted_peninsula}
ggplot(county_MI) + 
  geom_point(aes(x=Income, y=Black, colour=peninsula, size=TotalPop)) +
  ggtitle("Correlation of Blacks and Income by Peninsula") +
  facet_wrap(~peninsula)

ggplot(county_MI) + 
  geom_point(aes(x=Unemployment, y=Black, colour=peninsula, size=TotalPop)) +
  ggtitle("Correlation of Blacks and Unemployment by Peninsula") +
  facet_wrap(~peninsula)

ggplot(county_MI) + 
  geom_point(aes(x=SelfEmployed, y=Black, colour=peninsula, size=TotalPop)) +
  ggtitle("Correlation of Blacks and Self Employment by Peninsula") +
  facet_wrap(~peninsula)

```
*Correlation of Blacks and Income by Peninsula*
Well it seems as in counties with about 3% of African Americans have an average income of 45,000 and have a total population of 500,000 or less in the Lower Peninsula. One of the counties is an outlier with about 40% of African Americans and an average income of 45,000 and a Total Population of at least 1.5 million in the Lower Peninsula. In the Upper Peninsula, there is only one county, which is an outlier that has about 18% of African Americans. The rest of the counties have below 10% of Blacks. Each of the counties in the Upper Peninsula have less than 500,000 as Total Population and there is less than 10% of African Americans in these counties. There is one outlier county with about 18% of African Americans. In the Upper Peninsula, it looks like the average unemployment rate for African Americans is about 9%.

*Correlation of Blacks and Unemployment by Peninsula*
Well it seems as in counties with a higher population (1 million people or more) have about at least 10% of African Americans in the Lower Peninsula. The average unemployment rate for African Americans per county in the Lower Peninsula looks to be around 9%. There is an outlier county of a total population of 1.5 million with 40% of African Americans. They make up 16% of the unemployment rate in that county in the Lower Peninsula. In the Upper Peninsula, there are no counties with Total Poulations more than 500,000. is an outlier with about 40% of African Americans and an average income of 45,000 and a Total Population of at least 1.5 million in the Lower Peninsula. In the Upper Peninsula, there is only one county, which is an outlier that has about 18% of African Americans. The rest of the counties have below 10%. Each of the counties in the Upper Peninsula have less than 500,000 as Total Population and the average income for African Americans in each county in the Upper Peninsula is around 40,000.

*Correlation of Blacks and Self Employment by Peninsula*
Well it seems as in the points in this plot are skewed to the left and up for the percentage of African Americans, per county that are Self Employed in the Lower Peninsula. Meaning that the counties with the highest population as well as pergentage of blacks have tend to have very low percentages of blacks that are Self Employed. In the Upper Peninsula the points in this plot are somewhat similar to the Lower Peninsula. There is a light skew to the left and up.

Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}
ggplot(county_df) + 
  geom_point(aes(x=Income, y=TotalPop)) +
  ggtitle("Correlation of Total Population by Income for the USA")

```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
*Write a short paragraph on why one might want to use a log scale or logtransform of data when plotting.*
One may user log scale when looking at percent change is more important than the
actual change in value. One may also use log scale when data is highly skewed which can cause a majority of the values to be compressed into one small portion of your plot. Also, one may log transform data to try to change a non-linear relationship into a linear relationship

```{r}
ggplot(county_df) + 
  geom_point(aes(x=log(Income), y=TotalPop)) +
  ggtitle("Correlation of Total Population by Income for the USA")

```


### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.